---
title: "Gov 50 Final Project"
author: Paige Breyer
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---
## Milestone 1
I am interested in exploring data related to...
criminal justice. Specifically, I hope to look at data from states with the death penalty and those without to see whether or not capital punishment lowers crime rates. I am interested in learning how to create an accurate data set for this project, given that controlled experiments are more difficult in the social sciences. 

## Milestone 2
My research question for the final project is: do areas with a higher level of economic inequality experience more hate crimes? Before examining the evidence, I hypothesize that areas with more economic inequality will have a higher level of hate crimes due to the tension between communities that is often exacerbated by poverty. Individuals living in areas where financial hardship is more common are presumably more likely to feel animosity toward other groups of people, as competition for scarce resources is more acute. In areas with a higher level of economic inequality, there will likely be a higher level of hate crimes. In areas with more economic equality, there will likely be fewer hate crimes. My data sample (found in a code chunk in my R Markdown file) groups the variables (which include the gini index, poverty rates by race, etc.) by state. The explanatory variable for my experiment is the gini coefficient, which measures economic inequality on a scale from 0 to 1. The outcome variable is the number of hate crimes per 100,000 people in a given area. The data that I am using was compiled by FiveThrityEight and collected by both the FBI and the Southern Poverty Law Center. A positive correlation coefficient for the relationship between the gini index and hate crimes per capita would prove my hypothesis, while a negative correlation coefficient would disprove my hypothesis. 

## Milestone 3 
```{r}
library(readr)
hate_crimes <- read_csv("~/Downloads/hate-crimes/hate_crimes.csv") 
hate_crimes
```

```{r}
library(ggplot2)
hc_by_gini <- hate_crimes |> 
  ggplot(mapping = aes(x = gini_index, y = avg_hatecrimes_per_100k_fbi)) + 
  geom_point(color = "olivedrab") +
  geom_smooth(color = "chocolate4") + 
  labs(x = "Gini Index", 
       y = "Average Hate Crimes per Capita (FBI Data)")
hc_by_gini
```

```{r}
library(ggplot2)
hc_by_gini_2 <- hate_crimes |> 
  ggplot(mapping = aes(x = gini_index, y = hate_crimes_per_100k_splc)) + 
  geom_point(color = "forestgreen") +
  geom_smooth(color = "lavender") + 
  labs(x = "Gini Index", 
       y = "Average Hate Crimes per Capita (SPLC Data)")
hc_by_gini_2
```

## Milestone 4
```{r}
library(tidyverse)
viz_3 <- hate_crimes |> 
  drop_na() |> 
  summarize(cor(avg_hatecrimes_per_100k_fbi, gini_index))
knitr::kable(viz_3, col.names = c("Correlation Between Gini Index and Hate Crimes Per Capita"))
```
```{r}
library(tidyverse)
viz_4 <- hate_crimes |> 
  drop_na() |> 
  summarize(cor(hate_crimes_per_100k_splc, gini_index))
knitr::kable(viz_4, col.names = c("Correlation Between Gini Index and Hate Crimes Per Capita"))
```

The first four visuals (displayed above) display a positive correlation between economic inequality and hate crimes per capita in this data set. The correlation is positive for both the SPLC estimates of hate crimes per capita and the FBI estimates of hate crimes per capita. 

```{r}
hate_crimes <- hate_crimes |> 
  mutate(income_group = case_when(
    median_household_income < 50000 ~ "1. Under $50k", 
    median_household_income > 50000 & median_household_income < 60000 ~ "2. $50k - $60k",
    median_household_income > 60000 ~ "3. Over $60k")) 
```

```{r}
viz_5 <- hate_crimes |> 
  ggplot(mapping = aes(x = hate_crimes_per_100k_splc, y = income_group)) + 
  geom_col() + 
  labs(title = "Hate Crimes per Capita by Income (SPLC Data)", 
       x = "Hate Crimes per Capita (SPLC Data)", 
       y = "Income Group")
viz_5
```

```{r}
viz_6 <- hate_crimes |> 
  ggplot(mapping = aes(x = avg_hatecrimes_per_100k_fbi, y = income_group)) + 
  geom_col() + 
  labs(title = "Hate Crimes per Capita by Income (FBI Data)", 
       x = "Hate Crimes per Capita (FBI Data)", 
       y = "Income Group")
viz_6
```

As is demonstrated in the bar graph labeled "Hate Crimes per Capita by Income (SPLC Data)," the most hate crimes per capita occur in counties where the median household income is $50k-$60k. In the graph labeled "Hate Crimes per Capita by Income (FBI Data)," However, the most hate crimes are estimated to occur in counties with a median household income above $60k. 

```{r}
viz_7 <- hate_crimes |> 
  ggplot(mapping = aes(x = median_household_income, y = gini_index)) + 
  geom_point() + 
  labs(x = "Median Household Income", 
       y = "Gini Index", 
       title = "Income Inequality by Median Income")
viz_7
```

```{r}
viz_8 <- hate_crimes |>
  drop_na() |> 
  summarize(cor(median_household_income, gini_index))
knitr::kable(viz_8, col.names = c("Correlation Between Gini Index and Median Household Income"))
```

The two visuals above demonstrate that the relationship between median household income and income inequality is negative (that is, counties with a lower median income tend to have more economic inequality on average). 

```

